\emph{On Logics Extended with Embedding-Closed Quantifiers}

### 2020-11-21 paper.tex
Generalize the infinite case theorem to interpretations

\begin{lemma}\label{quant_elim_finite_lemma}
Let $\tau$ and $\sigma$ be finite relational vocabularies and
$(\Psi, (\psi_R)_{R \in \sigma})$ an
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-interpretation of
$\sigma$ in $\tau$.
Suppose that $Q$ is an embedding-closed quantifier of width $n < \omega$
and $\varphi = Q(\overline{x}_i \chi_i)_{i < n}$ a formula with $m$ free
variables where all $\chi_i$ are
quantifier-free $\sigma$-formulas.
Let $(\Psi(\mathfrak{A}_i))_{i < \omega}$ be a chain where all $\mathfrak{A}_i$
are quasi-homogeneous $\tau$-structures.
Then there is a natural number $N$ and a quantifier-free $\sigma$-formula
$\vartheta$ such that
\[
	\Psi(\mathfrak{A}_i) \vDash \forall
	\overline{x}(\varphi \leftrightarrow \vartheta)
\]
for all $i \geq N$.
\end{lemma}
\begin{proof}
For each $i < \omega$, let $T_i$ be the set of all atomic
$m$-types $t$ of $\tau$
such that $\mathfrak{A}_j, \overline{a} \vDash t$
and $\Psi(\mathfrak{A}_j), \overline{a} \vDash \varphi$
for some $\overline{a} \in A^m$ and $j \leq i$.
We clearly have $T_j \subseteq T_i$ whenever $j \leq i$.
Since $\tau$ is finite and relational, there is a natural number $k$
at which the chain $(T_i)_{i < \omega}$ achieves its maximum.
Write $T = T_k$.
We claim that for all $i \geq k$ and $\overline{a} \in A_i^m$, we have
\begin{equation}\label{eq_0}
\mathfrak{A}_i, \overline{a} \vDash \bigvee T
\Leftrightarrow
\Psi(\mathfrak{A}_i), \overline{a} \vDash \varphi.
\end{equation}
The statement on the right side implies the left side statement by the
definition of $T$.
For the other direction, suppose that
$\mathfrak{A}_i, \overline{a} \vDash \bigvee T$
and assume to the contrary that
$\Psi(\mathfrak{A}_i), \overline{a} \nvDash \varphi$.
By the definition of $T$, there is $j \leq i$ and $\overline{b} \in A_j^m$
such that $\overline{a}$ and $\overline{b}$ have the same atomic type of $\tau$
and $\Psi(\mathfrak{A}_j), \overline{b} \vDash \varphi$.
Since $\Psi(\mathfrak{A}_j) \leq \Psi(\mathfrak{A}_i)$, $Q$ is embedding-closed
and all $\chi_i$ are quantifier-free, we have
$\Psi(\mathfrak{A}_i), \overline{b} \vDash \varphi$.
However, since $\mathfrak{A}_i$ is quasi-homogeneous,
there is, by Lemma 2.2 and Theorem 5.3, a quantifier-free $\tau$-formula
$\varphi^*$ such that
$\mathfrak{A}_i, \overline{a} \nvDash \varphi^*$
and $\mathfrak{A}_i, \overline{b} \vDash \varphi^*$
which is a contradiction since $\overline{a}$ and $\overline{b}$ have the
same atomic types in $\tau$.
The claim is thus proven.

For each natural number $i \geq k$,
let $T'_i$ be the set of all atomic types $t$ of $\sigma$
for which there is a natural number $j$ with
$k \leq j \leq i$ and $\overline{a} \in A^m_j$ such that
$\Psi(\mathfrak{A}_j), \overline{a} \vDash t$ and
$\mathfrak{A}_j, \overline{a} \vDash \bigvee T$.
The chain $(T'_i)_{i < \omega}$ is clearly monotone and, since $\sigma$ is
finite and relational, there is a natural
number $k'$ at which the chain $(T'_i)_{i < \omega}$ reaches
its maximum.
Denote $T' = T'_{k'}$.
We claim that
\begin{equation}\label{eq_1}
\Psi(\mathfrak{A}_i), \overline{a} \vDash \bigvee T' \Leftrightarrow
\mathfrak{A}_i, \overline{a} \vDash \bigvee T
\end{equation}
for all $i \geq k'$ and $\overline{a} \in A_i^m$.

@@

\begin{theorem}\label{infinite}
Let $\tau$ be a vocabulary, $\kappa$ a cardinal, $\mathcal{Q}$ a set of
embedding-closed quantifiers of width less than $\kappa$, and $\lambda$ a regular
cardinal such that
\[
	2^{|\tau|\cdot\aleph_0} \cdot |\mathcal{Q}|^{\kappa} < \lambda.
\]
Suppose $(\mathfrak{A}_{\alpha})_{\alpha<\lambda}$ is a chain of quasi-homogeneous
$\tau$-structures.
There is a cardinal $\mu<\lambda$ such that for every formula
$\varphi \in \mathcal{L}_{\infty\omega}(\mathcal{Q})$ there is a quantifier-free
$\tau$-formula $\vartheta_{\varphi}$ such that
\[
	\mathfrak{A}_{\alpha} \vDash \forall \overline{x}(\varphi 
	\leftrightarrow \vartheta_{\varphi})
\]
for all $\alpha \geq \mu$.
In particular,
$\mathfrak{A}_{\alpha} \preceq_{\mathcal{L}_{\infty\omega}(\mathcal{Q})} \mathfrak{A}_{\beta}$
for all $\alpha$, $\beta$ with $\mu \leq \alpha \leq \beta < \lambda$.
\end{theorem}
\begin{proof}
Let $Q \in \mathcal{Q}$ and $\varphi = Q(\overline{x}_i\vartheta_i)_{i<\delta}$
where all $\vartheta_i$ are quantifier-free $\tau$-formulas.
In the first part of this proof, we will generalize the proof of Lemma
\ref{quant_elim_finite_lemma} so it works for chains of quasi-homogeneous
structures of arbitrary length.
First we observe that for all $\alpha< \lambda$ there is the smallest set
$T_{\alpha}$ of atomic types of $\tau$ such that
$\mathfrak{A}_{\alpha} \vDash \forall\overline{x}(\varphi \leftrightarrow \bigvee T_{\alpha})$
because of quasi-homogeneity of $\mathfrak{A}_{\alpha}$.
In addition, if $\alpha < \beta < \lambda$, then
$T_{\alpha} \subseteq T_{\beta}$.
Thus, since there are at most $2^{|\tau|+\aleph_0}$ atomic types of $\tau$,
if $\lambda$ is a regular cardinal greater than $2^{|\tau|+\aleph_0}$
and $(\mathfrak{A}_{\alpha})_{\alpha<\lambda}$ is a chain of quasi-homogeneous
$\tau$-structures, then there is a cardinal $\kappa_{\varphi}<\lambda$ and a
quantifier-free $\tau$-formula $\vartheta_{\varphi}$ that is equivalent to
$\varphi$ in structures $\mathfrak{A}_{\alpha}$ with $\alpha\geq\kappa_{\varphi}$.

Let $\Phi$ be the set of all formulas of the form
$Q(\overline{x}_i\vartheta_i)_{i<\delta}$,
with $Q\in\mathcal{Q}$ and all $\vartheta_i$ quantifier-free.
Then $|\Phi| \leq 2^{|\tau|\cdot\aleph_0}\cdot |\mathcal{Q}|^{\kappa}$.
We claim that we can set
$\mu := \sup\{\kappa_{\varphi} \colon \varphi \in \Phi \}$.
Note that $\mu<\lambda$.
We use induction on the structure of the formula to prove this claim.
If $\varphi$ is atomic there is nothing to prove.
If $\varphi = \bigwedge_{i\in I} \varphi_i$ and the claim holds for all
$\varphi_i$, then $\vartheta_{\varphi} = \bigwedge_{i\in I}\vartheta_{\varphi_i}$.
If $\varphi = \neg \varphi'$ and the claim is true for $\varphi'$,
then $\vartheta_{\varphi} = \neg \vartheta_{\varphi'}$.
Finally, suppose that $\varphi = Q(\overline{x}_i\varphi_i)_{i<\delta}$ and the
claim holds for all $\varphi_i$.
Then each $\varphi_i$ is equivalent to a quantifier-free formula $\vartheta_i$
on structures $\mathfrak{A}_{\alpha}$ with $\alpha \geq \mu$ so by the first part
of the proof, $\varphi$ is also equivalent to some quantifier-free formula
$\vartheta_{\varphi}$ on these structures.
\end{proof}

>>>>

\begin{lemma}\label{quant_elim_finite_lemma}
Let $\tau$ and $\sigma$ be finite relational vocabularies and
$(\Psi, (\psi_R)_{R \in \sigma})$ an
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-interpretation of
$\sigma$ in $\tau$.
Suppose that $Q$ is an embedding-closed quantifier of width $n < \omega$
and $\varphi = Q(\overline{x}_i \chi_i)_{i < n}$ a formula with $m$ free
variables where all $\chi_i$ are
quantifier-free $\sigma$-formulas.
Let $(\Psi(\mathfrak{A}_i))_{i < \omega}$ be a chain where all $\mathfrak{A}_i$
are quasi-homogeneous $\tau$-structures.
Then there is a natural number $N$ and a quantifier-free $\sigma$-formula
$\vartheta$ such that
\[
	\Psi(\mathfrak{A}_i) \vDash \forall
	\overline{x}(\varphi \leftrightarrow \vartheta)
\]
for all $i \geq N$.
\end{lemma}
\begin{proof}
For each $i < \omega$, let $T_i$ be the set of all atomic
$m$-types $t$ of $\tau$
such that $\mathfrak{A}_j, \overline{a} \vDash t$
and $\Psi(\mathfrak{A}_j), \overline{a} \vDash \varphi$
for some $\overline{a} \in A^m$ and $j \leq i$.
We clearly have $T_j \subseteq T_i$ whenever $j \leq i$.
Since $\tau$ is finite and relational, there is a natural number $k$
at which the chain $(T_i)_{i < \omega}$ achieves its maximum.
Write $T = T_k$.
We claim that for all $i \geq k$ and $\overline{a} \in A_i^m$, we have
\begin{equation}\label{eq_0}
\mathfrak{A}_i, \overline{a} \vDash \bigvee T
\Leftrightarrow
\Psi(\mathfrak{A}_i), \overline{a} \vDash \varphi.
\end{equation}
The right hand side implies the left hand side by the
definition of $T$.
For the other direction, suppose that
$\mathfrak{A}_i, \overline{a} \vDash \bigvee T$
and assume to the contrary that
$\Psi(\mathfrak{A}_i), \overline{a} \nvDash \varphi$.
By the definition of $T$, there is $j \leq i$ and $\overline{b} \in A_j^m$
such that $\overline{a}$ and $\overline{b}$ have the same atomic type of $\tau$
and $\Psi(\mathfrak{A}_j), \overline{b} \vDash \varphi$.
Since $\Psi(\mathfrak{A}_j) \leq \Psi(\mathfrak{A}_i)$, $Q$ is embedding-closed
and all $\chi_i$ are quantifier-free, we have
$\Psi(\mathfrak{A}_i), \overline{b} \vDash \varphi$.
However, since $\mathfrak{A}_i$ is quasi-homogeneous,
there is, by Lemma \ref{interpr} and Theorem \ref{homog}, 
a quantifier-free $\tau$-formula
$\varphi^*$ such that
$\mathfrak{A}_i, \overline{a} \nvDash \varphi^*$
and $\mathfrak{A}_i, \overline{b} \vDash \varphi^*$
which is a contradiction since $\overline{a}$ and $\overline{b}$ have the
same atomic types in $\tau$.
The claim is thus proven.

For each natural number $i \geq k$,
let $T'_i$ be the set of all atomic types $t$ of $\sigma$
for which there is a natural number $j$ with
$k \leq j \leq i$ and $\overline{a} \in A^m_j$ such that
$\Psi(\mathfrak{A}_j), \overline{a} \vDash t$ and
$\mathfrak{A}_j, \overline{a} \vDash \bigvee T$.
The chain $(T'_i)_{k \leq i < \omega}$ is clearly monotone and, 
since $\sigma$ is finite and relational, 
there is a natural number $k'$ at which it reaches its maximum.
Denote $T' = T'_{k'}$.
We claim that
\begin{equation}\label{eq_1}
\Psi(\mathfrak{A}_i), \overline{a} \vDash \bigvee T' \Leftrightarrow
\mathfrak{A}_i, \overline{a} \vDash \bigvee T
\end{equation}
for all $i \geq k'$ and $\overline{a} \in A_i^m$.

@@

\begin{theorem}\label{infinite}
Let $\tau$ and $\sigma$ be relational vocabularies, $\kappa$ a cardinal, 
$\mathcal{Q}$ a set of embedding-closed quantifiers of width less than $\kappa$, 
and $\lambda$ a regular cardinal such that
\[
	2^{(|\tau| + |\sigma|) \cdot \aleph_0} \cdot |\mathcal{Q}|^{\kappa} < \lambda.
\]
Let $(\Psi, (\psi_R)_{R \in \sigma})$ be an
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-interpretation of
$\sigma$ in $\tau$, and $(\Psi(\mathfrak{A}_{\alpha}))_{\alpha < \lambda}$
a chain in which all $\mathfrak{A}_{\alpha}$ are quasi-homogeneous 
$\tau$-structures.
There is a cardinal $\mu<\lambda$ such that for every formula
$\varphi \in \mathcal{L}_{\infty\omega}(\mathcal{Q})[\sigma]$ 
there is a quantifier-free
$\sigma$-formula $\vartheta_{\varphi}$ such that
\[
	\mathfrak{A}_{\alpha} \vDash \forall \overline{x}(\varphi 
	\leftrightarrow \vartheta_{\varphi})
\]
for all $\alpha \geq \mu$.
In particular,
$\mathfrak{A}_{\alpha} \preceq_{\mathcal{L}_{\infty\omega}(\mathcal{Q})} 
\mathfrak{A}_{\beta}$
for all $\alpha$, $\beta$ with $\mu \leq \alpha \leq \beta < \lambda$.
\end{theorem}
\begin{proof}
Let $Q \in \mathcal{Q}$ and $\varphi = Q(\overline{x}_i\vartheta_i)_{i<\delta}$
where all $\vartheta_i$ are quantifier-free $\tau$-formulas.
In the first part of this proof, we will generalize the proof of Lemma
\ref{quant_elim_finite_lemma} so it works for chains of quasi-homogeneous
structures of arbitrary length.

For each $\alpha < \lambda$, let $T_{\alpha}$ be the set of all atomic 
types of $\tau$ such that $\mathfrak{A}_{\alpha}, \overline{a} \vDash t$
and $\Psi(\mathfrak{A}_{\alpha}), \overline{a} \vDash \varphi$ 
for some $\overline{a}$.
Since there are at most $2^{|\tau|+\aleph_0}$ atomic types of $\tau$,
and $\lambda$ is a regular cardinal greater than $2^{|\tau|+\aleph_0}$,
there is a cardinal $\kappa_{\varphi} < \lambda$ at which the chain 
$(T_{\alpha})_{\alpha < \lambda}$ reaches its maximum.
We can show in the same way as in the proof of 
Lemma \ref{quant_elim_finite_lemma} that for all 
$\alpha \geq \kappa_{\varphi}$ and 
$\overline{a} \in A^{\frvar(\varphi)}$ we then have
\[
	\mathfrak{A}_{\alpha}, \overline{a} \vDash \bigvee T_{\kappa_{\varphi}}
	\Leftrightarrow
	\Psi(\mathfrak{A}_{\alpha}), \overline{a} \vDash \varphi.
\]

For each $\alpha \geq \kappa_{\varphi}$, $\alpha < \lambda$, let $T_{\alpha}'$
be the set of all atomic types $t$ of $\sigma$ such that there is a cardinal 
$\beta$, $\kappa_{\varphi} \leq \beta \leq \alpha$, and
$\overline{a} \in A^{\frvar(\varphi)}_{\beta}$
for which
$\Psi(\mathfrak{A}_{\beta}), \overline{a} \vDash t$ and
$\mathfrak{A}_{\beta}, \overline{a} \vDash \bigvee T_{\kappa_{\varphi}}$.
The chain $(T'_{\alpha})_{\kappa_{\varphi} \leq \alpha < \lambda}$ achieves
its maximum at some cardinal $\mu_{\varphi} < \lambda$ that exists since
there are at most $2^{|\sigma| + \aleph_0}$ atomic types of $\sigma$
and $\lambda$ is a regular cardinal greater than this number.
Again, as in Lemma \ref{quant_elim_finite_lemma}, we can show that for all
$\alpha \geq \mu_{\varphi}$ and 
$\overline{a} \in A^{\frvar(\varphi)}_{\alpha}$ we have
\[
	\Psi(\mathfrak{A}_{\alpha}), \overline{a} \vDash \bigvee T_{\mu_{\varphi}}'
	\Leftrightarrow \mathfrak{A}_{\alpha}, \overline{a} \vDash \bigvee T_{\kappa_{\varphi}}
\]
and further
\[
	\Psi(\mathfrak{A}_{\alpha}), \overline{a} \vDash \bigvee T_{\mu_{\varphi}}'
	\Leftrightarrow \Psi(\mathfrak{A}_{\alpha}), \overline{a} \vDash \varphi.
\]
Thus, for every formula $\varphi$ of the form 
$Q(\overline{x}_i \vartheta_i)_{i < \delta}$ with all $\vartheta_i$
quantifier-free there is a cardinal $\mu_{\varphi} < \lambda$ and a 
quantifier-free formula $\vartheta_{\varphi}$ such that
\[
	\Psi(\mathfrak{A}_{\alpha}) \vDash 
	\forall \overline{x}(\varphi \leftrightarrow \vartheta_{\varphi})
\]
for all $\alpha \geq \mu_{\varphi}$, $\alpha < \lambda$.

Let $\Phi$ be the set of all $\sigma$-formulas of the form
$Q(\overline{x}_i\vartheta_i)_{i<\delta}$,
with $Q\in\mathcal{Q}$ and all $\vartheta_i$ quantifier-free.
Then 
$|\Phi| \leq 2^{(|\sigma| + |\tau|) \cdot \aleph_0} \cdot |\mathcal{Q}|^{\kappa}$.
We claim that we can set
$\mu := \sup\{\mu_{\varphi} \colon \varphi \in \Phi \}$.
Note that $\mu < \lambda$.
We use induction on the structure of the formula to prove this claim.
If $\varphi$ is atomic there is nothing to prove.
If $\varphi = \bigwedge_{i \in I} \varphi_i$ and the claim holds for all
$\varphi_i$, then 
$\vartheta_{\varphi} = \bigwedge_{i\in I}\vartheta_{\varphi_i}$.
If $\varphi = \neg \varphi'$ and the claim is true for $\varphi'$,
then $\vartheta_{\varphi} = \neg \vartheta_{\varphi'}$.
Finally, suppose that $\varphi = Q(\overline{x}_i \varphi_i)_{i < \delta}$ 
and the claim holds for all $\varphi_i$.
Then each $\varphi_i$ is equivalent to a quantifier-free formula $\vartheta_i$
on structures $\mathfrak{A}_{\alpha}$ with $\alpha \geq \mu$ so by the first part
of the proof, $\varphi$ is also equivalent to some quantifier-free formula
$\vartheta_{\varphi}$ on these structures.
\end{proof}
